---
title: Replication of the article "Emerging Parental Gender Indifference? Sex Composition
  of Children and the Third Birth" From Pollard and Morgan
author: "Juliana de Oliveira Horst"
date: "May 11, 2022"
output:
  pdf_document:
    keep_tex: yes
    fig_caption: yes
  word_document: default
  html_document:
    df_print: paged
geometry: left=3cm,right=3cm,top=2cm,bottom=2cm
fontfamily: mathpazo
fontsize: 12pt
urlcolor: black
header-includes:
- \pagenumbering{arabic}
- \usepackage{setspace}\doublespacing
---


```{r setup, include=FALSE}
# This cell contains default chunk options
# These will be applied to all chunks unless an individual chunk is modified
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(dev = 'pdf')
# You may modify this chunk to include additional options
# For example, there are options that can change the layout for all figures
# You can also modify the header options above this chunk to change the
# formatting of the output, but do so at your own risk.
# This book provides lots of tips for using RMarkdown: https://bookdown.org/yihui/rmarkdown-cookbook/
```


# INTRODUCTION

  In “Emerging Parental Gender Indifference? Sex Composition 
  of Children and the Third Birth,” Pollard and Morgan’s 
  central hypothesis is that there has been a transformation
  in parents’ preferences regarding the sex of their offspring. 
  According to the authors, because of ruptures in the “gender
  system” – represented, in this article, mainly by the 
  gendered division of labor –, parents’ desire to have 
  at least one child of each sex is declining in the US. This 
  hypothesis opposes well-established literature pointing 
  out that parents of two same-sex children are more likely
  to have (or intend to have) a third child than those that
  already have one child of each sex (Slone and Lee, 1983; 
  Yamaguchi and Ferguson, 1995). To build their arguments,
  the authors developed some Logistic Regressions based on 
  varying data sources, especially four cycles from the 
  Current Population Survey (1980,1985, 1990, and 1995) 
  and three cycles from the National Survey of Family Growth 
  (1983, 1988, and 1995). 
  
  Their models – which had “fertility behavior” and “fertility
  intention” as their main independent variables– indeed 
  present a significant decline, after 1985, in the likelihood 
  of mothers of same-sex children having (fertility behavior) 
  or intending to have (fertility intention) a third child. 
  The authors aim to attribute this transformation to the 
  aforementioned “weakening of the gender system” throughout 
  the article. To corroborate this explanation, they attempt to
  falsify other plausible justifications for the phenomenon. 
  One critical competing hypothesis for the current decline is 
  the changes in families’ social structure, particularly the 
  increasing number of single-parent households. 
  
  To verify whether the increase in “indifference” toward 
  the sex of children is related to “non-traditional” family
  structures or whether it is pervasive, the authors created
  a more restricted logistic regression model. Model number 
  three, which I will replicate in this paper, has only two 
  categories of respondents: “(1) women who were married at 
  the birth of their second child, and (2) once-married women 
  who were married at the birth of the first child and remained
  married until the third birth or censoring” (p. 607). Thus,
  using the author’s rationale, if a decrease in the influence 
  of the sex of the first two children on women's intention to 
  have a third child could be detected in this group, it would 
  mean the phenomenon is not only linked to non-traditional families.
  
  Similar to models one and two, based on data from a larger
  population, model three also demonstrated a significant decline
  in the influence of the sex of the first two children on the 
	intention to have a third child. Thus, the authors' hypothesis 
	became even more robust. I will replicate and analyze model 
	three, its assumptions, and findings in the following sections.
	


# REPLICATION

  The authors have made the codes they developed to create 
  their logistic models widely accessible to the public. As such,
  the model I replicated is reasonably close to theirs. In the
  original model three, the authors first created five subsets 
  from their more extensive database, keeping only the married 
  women population and their corresponding demographic data. 
  The five subsets - which they also grouped into a new dataset 
  (married.ASR) - correspond to the different years the surveys 
  were conducted, 1982, 1988, 1995, 2002, and 2006.
  
  According to my replication, one can notice that the variable 
  “same-sex children” (“samesexsame-sex”) still has a significant
  influence on married women’s intention to have a third child (0.48),
  albeit declining. Two other statistically relevant figures 
  concern the variable “age,” where increasing age significantly 
  decreases the prospects of a third child (-0.21), and the 
  variable “raceHispanics.” In the latter, there is a rather 
  significant influence of the gender of the first two children 
  on the intention to have a third child (0.77), which can 
  suggest a great ethnic-cultural difference between the Hispanic 
  group and the others.
  
  Finally, although less salient, it is also important to
  mention that there appears to be a positive relationship 
	between the variable “educatcollege” with the intention of 
	married women to have a third child. The data suggests that 
	the gender of the first two children tends to influence (more) 
	the will of women who went to college to have a third child 
	(0.42). This information is relevant because, throughout the 
	article, Pollard and Morgan argue that Yamaguchi and 
	Ferguson's (1995) findings – that the influence of the 
	sex of previous children is more substantial in highly 
	educated women – should not prosper (Pollard and Morgan,
	p. 607, 2002). However, looking at Pollard and Morgan’s 
	own data, we need more caution before ruling out this hypothesis. 
	

```{r data loading, include=FALSE, echo=TRUE, tidy=TRUE}
rm(list=ls(all=TRUE))

library(foreign)
library(MASS)
library(RColorBrewer)
library(tidyverse)
library(rstanarm)
library(modelsummary)
library(formatR)
options(tinytex.verbose = TRUE)

# Load data
wd<-c("C:/Users/julia/OneDrive/Área de Trabalho/Nova pasta")
setwd(wd)
data = read.dta("intent.dta")
ASR = data[which(data$survey==1988 | data$survey==1995),]
NSFG1982 = data[which(data$survey==1982),]
NSFG1988 = data[which(data$survey==1988),]   
NSFG1995 = data[which(data$survey==1995),]
NSFG2002 = data[which(data$survey==2002),]
NSFG2006 = data[which(data$survey==2006),]
```

```{r replication of the original model, include=TRUE, echo=TRUE, tidy=TRUE}
# m3: currently married sample:  + demographic controls
  married.ASR = ASR[which(ASR$marital
                          =="currently married"),]
  married.1988 = NSFG1988[which(NSFG1988$marital
                                =="currently married"),]
  married.1995 = NSFG1995[which(NSFG1995$marital
                                =="currently married"),]
  married.1982 = NSFG1982[which(NSFG1982$marital
                                =="currently married"),]
  married.2002 = NSFG2002[which(NSFG2002$marital
                                =="currently married"),]
  married.2006 = NSFG2006[which(NSFG2006$marital
                                =="currently married"),]
  m3 = glm(intent ~ samesex+as.factor(survey)+
             samesex*as.factor(survey)+age+educat+race, 
           data=married.ASR,family=binomial(link="logit"))
  m3.1988 = glm(intent ~ samesex+age+educat+race, 
                data=married.1988, 
                family=binomial(link="logit"))
  m3.1995 = glm(intent ~ samesex+age+educat+race, 
                data=married.1995,
                family=binomial(link="logit"))
  m3.1982 = glm(intent ~ samesex+age+educat+race, 
                data=married.1982, 
                family=binomial(link="logit"))
  m3.2002 = glm(intent ~ samesex+age+educat+race, 
                data=married.2002, 
                family=binomial(link="logit"))
  m3.2006 = glm(intent ~ samesex+age+educat+race, 
                data=married.2006, 
                family=binomial(link="logit"))
    coef.m3 = round(c(coef(m3.1982)[2],
                      coef(m3.1988)[2],coef(m3.1995)[2],
                      coef(m3.2002)[2],coef(m3.2006)[2]), 
                    digits=3)
    se.m3 = round(c(sqrt(diag(vcov(m3.1982)))[2],
                  sqrt(diag(vcov(m3.1988)))[2],
                  sqrt(diag(vcov(m3.1995)))
                  [2],sqrt(diag(vcov(m3.2002)))[2],
                  sqrt(diag(vcov(m3.2006)))[2]), 
                  digits=3)
    n.m3 = c(m3.1982$df.null+1, m3.1988$df.null+1, 
             m3.1995$df.null+1,
             m3.2002$df.null+1, m3.2006$df.null+1)
    sig.m3 = round(c(summary(m3.1982)$coef[2,4],
                  summary(m3.1988)$coef[2,4],
                  summary(m3.1995)$coef[2,4],
                  summary(m3.2002)$coef[2,4],
                  summary(m3.2006)$coef[2,4]),
                  digits=3)
    summary(m3)
    modelsummary(list("Initial model" = m3), 
             statistic = "conf.int",
             title = "Original replication")


```
# BAYESIAN REPLICATION

  As we can notice, the Bayesian replication presents results
  that are very similar to the original logistic regression. 
  The most extensive variation is in the variable “raceother.”
  In the original model, the value was 0.65, while in the Bayesian
  replication, this value is 0.57. The higher sensitivity to
  variation may be linked to a reduced sample size for this 
  category. However, the other variables present almost 
  identical results in both models, corroborating how 
  robust the authors’ findings are.
  

```{r bayesian replication, include=TRUE, echo=TRUE, tidy=TRUE}
#Bayesian model for m3
baym3 <- stan_glm(intent ~ samesex+as.factor(survey)+samesex*as.factor(survey)
                  +age+educat+race, data=married.ASR,
                  family=binomial(link="logit"),
                  prior = normal(location = 0,scale = 1), 
                  seed = 12345, refresh = 0)
print(baym3)
modelsummary(list("Initial model" = m3,
                  "Bayesian model" = baym3), 
             statistic = "conf.int",
             title = "Bayesian Replication for Model 3")
```

```{r posterior predictive check, include=TRUE, echo=TRUE, tidy=TRUE}
pp_check(baym3)

```
# ALTERNATIVE SPECIFICATIONS

  Another way to check how solid the results Pollard and 
  Morgan obtained were, is to employ sensitivity checks. 
  Therefore, in this section I propose three alternative 
  specifications for the replicated model; in the first 
  one I added a new variable (“catholic”), in the second 
  I altered the original dataset removing age outliers, and 
  in the last one, I incorporated the data into a Probit 
  regression model.  In the first specification, I decided 
  to add the variable “catholic” because this population 
  tends to be more prone to accepting gender norms. So, it 
  would be interesting to see how catholic married women 
  would perform. As we can notice, after adding this new 
  variable, most results did not alter significantly compared 
  to the initial one. This pattern corroborates the authors’ 
  argument that there is a widespread change regarding parental
  preference for the sex of their children, even when we observe
  a group that tends to be more impacted by the socialization and
  institutionalization of gender roles. 
  
  The only significant 
  variations are caused by the interaction between the 
  variable “catholic” and the	variables “raceHispanic” and 
  “raceblack.” The values went from 0.78 (“raceHispanic”) 
  and 0.06 (“raceblack”) in the	initial specification to 0.5 
  and 0.18 with the new variable. These discrepancies may 
  suggest either that there is indeed a sui generis phenomenon 
  occurring in these groups or that the sample size was small,
  rendering these categories more susceptible to unrealistic 
  variation.	
  
  In the second alternative specification, when I altered the 
  original dataset by removing age outliers, the  results found	
  were also very similar to the initial ones. With  the exception,
  once again, of categories involving race (now  also including
  “raceother”). I believe that this consistent discrepancy involving
  the non-white racial groups points a 
  lot more to issues in collecting this data than to any other 
  possible underlying explanation. In that sense, Pollard 
  and Morgan also state that they were “not able to reject 
  the hypothesis that the sex-of-previous-children effect was 
  the same across racial/ethnic groups (results not shown)” 
  (p. 609), this also corroborates my assumption that there 
  is a lack of comprehensive data associated with the racial 
  categories.	
  
  Finally, the third scenario was utilizing an 
  alternative model. Differently from the other two specifications, 
  the Probit regression model displayed results that are very 
  dissimilar when compared to the initial logistic model. However,
  we can notice an important pattern between the two models. 
  There is a certain proportion between the new values and the
  previous ones, for the positive outcomes arising from Probit 
  are (roughly) half of Logit’s outcomes. For instance, when we 
  observe the variable “educatcollege” it goes from 0.423 in the 
  Logit to 0.225 in the	Probit; “samesexsame-sex” goes from 0.483 
  to 0.245; “raceother” goes from 0.657 to 0.375. This pattern
  might suggest that thefundamental relationships between the
  variables are maintained in the new model, even if the values 
  are not. However, it is essential to get more information to 
  understand further what happened in this part of the 
  sensitivity checks	mainly because the general expectation is 
  that Logit and Probit’s results usually	are similar.
	


```{r alternative specifications, include=TRUE, echo=TRUE, tidy=TRUE}
m3catholic = glm(intent ~ samesex+as.factor(survey)+samesex*as.factor(survey)
                 +age+educat+race+catholic, 
                 data=married.ASR,
                 family=binomial(link="logit")) 
m3probit = glm(intent ~ samesex+as.factor(survey)+samesex*as.factor(survey)
               +age+educat+race, 
               data=married.ASR,
               family=binomial(link="probit")) 
married.ASR <- married.ASR %>% 
  mutate(age = ifelse(age > 
                        quantile(age, .95, na.rm = T) | 
                        age < quantile(age, .05, na.rm = T), 
                      NA, age))
m3quantile = glm(intent ~ samesex+as.factor(survey)+
                   samesex*as.factor(survey)
                 +age+educat+race, data=married.ASR,
                 family=binomial(link="logit")) 
modelsummary(list("Initial model" = m3,
                  "Add Catholic" = m3catholic,
                  "Probit" = m3probit,
                  "Transformed data" = m3quantile),
             title = "Alternative Specifications")
```
# DISCUSSION

  The results Pollard and Morgan obtained in this research are 
  quite robust, remaining virtually unchanged both when transformed 
  into a Bayesian regression model and when subjected to alternative
  specifications (apart from the Probit model). So, from a statistical
  data production standpoint, their outcomes are well supported. 
  However, the authors’ analysis of the social phenomenon arising 
  from their results are less impressive and unconvincing. Even if 
  they are correct and, since 1985, couples in the United States
  have been increasingly indifferent to having at least one child
  of each sex. This finding is not synonymous with saying that 
  there is an increase in parents’ indifference concerning the sex of
  their children. 
  
  To be able to more forcefully state that there 
  is indeed an increase in parental indifference towards the sex 
  of their children, it would also be necessary to analyze whether 
  one sex is still more desired over the other. For example, what
  percentage of parents would like to have only girls or only
  boys, what percentage of parents who have their first child a
  girl decides to have a second child, and vice versa. Sticking 
  to the influence of the sex of the first two children, it would 
  be important to inquire how often parents of two girls decide 
  to have a third child, is it the same as parents of two boys? 
  
  Moreover, attributing the new trend they identified to the 
  attenuation of the “gender system” is rather one-dimensional. After
  all, gender norms have historically gone through shifts	without 
  it necessarily resulting from a weakening in this system.
	Even the advancement of women's rights – such as their incorporation
	into the labor market pointed out several times in the article
	– often goes hand in hand with increasing violence against them to
	reestablish and reinforce gender norms (Faludi, 2006). 
	
	To argue 	for some relevant weakening of the patriarchal system,
	it is necessary that, at the very least, women and men exhibited 
	changes in behaviors and intentions. However, the population
	Pollard and Morgan investigated were only composed by 
	women-mothers. Throughout the article, but especially when 
	analyzing the data stemming from model number three, the authors 
	assume a homology between the women’s intentions and those of 
	their husbands. This equation may be far from reality.
	
	Finally, it	was quite interesting to carry out this replication
	exercise, following the steps of others in producing their 
	statistical research. The most striking aspect for me was 
	visualizing how broad a quantitative researcher's decision-making 
	power is. From deciding which data they will incorporate or ignore,
	which categories they will create or dissolve, which clusters 
	they will group or separate, which statistical model is more 
	interesting to present their results, etc. And how all these 
	decisions, as small as they may seem, can significantly alter 
	their outcomes.
	


# REFERENCES

Faludi, Susan. 2006 [1991]. Backlash: The Undeclared War Against 
American Women. New York: Three Rivers Press. 

Pollard, Michael S. and S. Philip Morgan. 2002. “Emerging 
parental gender indifference? Sex composition of children and 
the third birth.” American Sociological Review 67: 600-613.

Sloane, Douglas M. and Che-Fu Lee. 1983. “Sex of previous children
and intentions for further births in the United States, 
1965-1976.” Demography 20: 353-367.

Yamaguchi, Kazuo and Linda R. Ferguson. 1995. “The stopping and 
spacing of childbirths and their birth-history predictors: 
rational-choice theory and event-history analysis.” American 
Sociological Review 60: 272-298.
